# Multiple Regression Review {#module-2}

## Learning Objectives

In this module, we will review simple and multiple linear regression to establish a strong foundation before moving onto multilevel models. Note that this is intended more as review than a comprehensive guide to regression; for the latter, we recommend https://stats.oarc.ucla.edu/other/mult-pkg/seminars/.

All materials for this chapter are available for download [here](#appendix).

The learning objectives for this chapter are:

1. Understand using file paths for project management/loading data.
2. Review using simple and multiple linear regression to analyze data.

## Data Demonstration

In this data demo, we will first review setting up an R session, then simple and multiple linear regression.

The data for this chapter were taken from chapter 3 of Heck, R. H., Thomas, S. L., & Tabata, L. N. (2011). *Multilevel and Longitudinal Modeling with IBM SPSS*: Taylor & Francis. These data have a multilevel structure, which we will work with in chapter 3, but for this chapter we will ignore the clustering structure and conduct regular regression. The following variables are in this data set:

|Variable|Level|Description|Values|Measurement|
|:-------|:----|:----------|:-----|:----------|
|schcode|School|School identifier (419 schools)|Integer|Ordinal|
|Rid|Individual|A within-group level identifier representing a sequential identifier for each student within 419 schools.|1 to 37|Ordinal|
|id|Individual|Student identifier (6,871 students)|Integer|Ordinal|
|female|Individual|Student sex|0 = Male, 1 = Female|Scale|
|ses|Individual|Z-score measuring student socioeconomic status composition within the schools|-2.41 to 1.87|Scale|
|femses|Individual|Grand-mean-centered variable measuring student socioeconomic status by gender (female)|-2.41 to 1.85|Scale|
|math|Individual|Student math achievement test score|27.42 to 99.98|Scale|
|ses_mean|School|Grand-mean-centered variable measuring student socioeconomic status|-1.30 to 1.44|Scale|
|pro4yrc|School|Aggregate proportion of students who intend to study at 4-year universities|0.00 to 1.00|Scale|
|public|School|Dichotomous variable identifying school type|0 = Other, 1 = Public School|Scale|

### Creating R Projects

Before we get into analyzing the data, let's start by creating a new project file for this module. R project files help you keep all of the files associated with your project -- data, R scripts, and output (including figures) -- in one location so you can easily navigate everything related to your project. 

To create a project, open R, click "File" and "New Project...". If you have already created a folder for this chapter, you can add an R Project to that folder by clicking "Existing Directory"; the R project file will take on the name of that folder. If you do not already have a folder, click "New Directory," choose where you want to put your new folder and what you want to call it. The R Project file will again take on the name of your new folder.

### Loading Data and Dependencies

Next, let's load in the data and packages we'll be using for this demo. We'll be using the following packages:

```{r data-dependencies, message=FALSE, warning=FALSE}
library(ggplot2) # for data visualization
library(magrittr) # for pipe, %>% 
```

You must install a given package before you can use it. For example: `install.package("ggplot2")`. Once you have installed a package, you can load it into any future sessions with `library(package_name)`.

Next, let's read in the data. If you have your code and data in the same directory, you can read the data in as follows:

```{r data, eval=FALSE}
data <- read.csv('heck2011.csv')
```

```{r, echo = FALSE}
# this actually loads my code, but will be hidden
data <- read.csv('data/heck2011.csv')
```

This is called a *relative* file path, because you're telling your computer where to find the data relative to your current folder (a folder can also be called a "directory"). You could also use an absolute file path that fully states where your files are located, like: 

```{r, eval=FALSE}
read.csv('/Users/maireadshaw/open_mlm_materials/heck2011.csv')
```

Let's calculate some descriptive statistics and compare them to the above table to make sure we read our data in correctly.

```{r}
summary(data)
```

That looks good, so let's proceed to conducting regressions.

### Simple Linear Regression

Let's run a simple linear regression predicting math achievement (`math`) from socioeconomic status (`ses`). The syntax for the `lm()` (linear modelling) command in R is `lm(DV ~ IV1 + IV2 + ... + IVn, data = dataframe)`.

```{r simple-regression}
model1 <- lm(math ~ ses, data = data)
summary(model1)
```

The intercept from this regression is 57.60, indicating that students at the mean level of SES within a school (i.e., when SES = 0, given that SES is z-scored) have an average math achievement score of 57.6 out of 100. This score is significantly different from 0, per the p-value. 

Per the coefficient for SES, a one-unit increase in SES is associated with a 4.25-point increase in student math achievement on average, also significant.

The adjusted R-squared value is 14.3%, indicating that 14.3% of the variance in math achievement is explained by socioeconomic status.

We can visualize this relationship by graphing a scatter plot.

```{r graph}
ggplot(data = data, mapping = aes(x = ses, y = math)) +
  geom_point()
```

Our graph reflects the positive relationship between SES and math achievement (and also shows a lot of math scores collecting around the 60 mark). 

### Multiple Regression

Next, let's add the available sex variable `female` (0 = male, 1 = female) as a predictor in our regression and interpret the coefficients and R-squared value.

```{r}
model2 <- lm(math ~ ses + female, data = data)
summary(model2)
```

The intercept of 58.13 reflects the average math achievement score (out of 100) for male students (`female` = 0) at their class average SES (`ses` = 0). For a one-unit increase in SES, math achievement increases by 4.23 points, controlling for sex. Female students had a math achievement score lower by 1.06 points on average, controlling for SES. SES and sex together explain 14.6% of the variance in math achievement.

### Interaction Terms

In the previous model, we assumed that the relationship between SES and math achievement was constant for both sexes (homogeneity of regression slopes, i.e., an ANCOVA model). As a final exercise, let's add an interaction term to our regression between sex and SES.

```{r}
model3 <- lm(math ~ ses + female + ses:female, data = data)
summary(model3)

# Could also succinctly code it as follows:
# lm(math ~ ses*female, data = data)
```

An interaction captures that the relationship between two variables may differ based on the level of another variable (i.e., different slopes for different folks). An interaction term, A:B, has two possible interpretations:

1. The effect of A on the effect of B on your outcome Y.
2. The effect of B on the effect of A on your outcome Y.

The `ses:female` interaction term, .34, represents the effect of being female on the relationship between SES and math achievement. Alternatively, it could represent the effect of SES on the relationship between being female and math achievement. In this case, the latter is a more intuitive interpretation: female students from higher socioeconomic statuses are slightly insulated from the negative relationship between female and math achievement in this sample. As SES increases by one point, the relationship between being female and math achievement becomes less negative, from -1.07 to -.73 (-1.07 + .34). However, this interaction term is not statistically significantly different from zero per the p-value.

We can see this graphically using the `sjPlot` package:

```{r}
sjPlot::plot_model(model3, type = "pred", terms = c("ses", "female"))
```

As we can see, math scores for males (`female` = 0, the red line) are higher than those for females (`female` = 1, the blue line) at all levels of SES. However, the difference between males and females shrinks with increasing SES, as indicated by math scores at higher SES levels being closer than those at lower levels of SES.

The other coefficients have the same interpretations as before. The R-squared indicates that SES and sex account for 14.7% of the variance in math achievement.

## Conclusion

If you feel comfortable with the material presented in this data demonstration, then you have a sufficiently strong baseline to move forward with the materials. In this chapter, we ignored that students were clustered into schools; in the next chapter, we'll examine that clustering, consider its implications for our analyses, and introduce one non-multilevel-model method for handling clustered data.